---
title: "Linear Model using Python"
description: |
  Linear Model using Python.
author:
  - name: Muhammad Yaseen 
    url: http://myaseen208.netlify.com/
    affiliation: Dept of Math & Stat, Univ of Agriculture, Faisalabad Pakistan
    affiliation_url: http://uaf.edu.pk/
date: 2019-02-18
comments: false
slug: 2019-PythonLMBZU
categories:
- Python
- Statistics
- Linear Model
- Regression Analysis
- ANOVA
- ANCOVA
mathjax: true
output:
  blogdown::html_page:
    toc: true
header-includes: 
  - \usepackage{tikz}
  - \usepackage{pgfplots}    
---

```{r, label = setup, include = FALSE, cache = FALSE}
library(knitr)
opts_chunk$set(
              #  fig.path   = 'figure/'
                cache.path = 'cache/'
               , cache      = FALSE
               , dev        = c("svg", "tikz", "pdf", "png")[1]
               , fig.align  = 'center'
              # , fig.show   = 'asis'
              # , fig.width  = 6
              # , fig.height = 6
              # , out.width  = '0.9\\linewidth'
               , par        = TRUE
               , echo       = TRUE
               , results    = 'markup'
               , warning    = FALSE
               , message    = FALSE
               , comment    = NA
               )
options(
          replace.assign = TRUE
        , width          = 80
        , knitr.kable.NA = ''
        )

knit_hooks$set(par = function(before, options, envir){
if (before && options$fig.show!='none')
 par(mar=c(4, 4, 0.1, 0.1), cex.lab = 0.95, cex.axis =0.9, mgp = c(2, 0.7, 0), tcl = -0.3)
}, crop = hook_pdfcrop)

# Suppressing Error Messages in knitr
muffleError <- function(x,options) {}
knit_hooks$set(error=muffleError)
```

```{r, echo = FALSE}
library(reticulate)
reticulate::use_python(python = Sys.which("python3"), required = TRUE)
# reticulate::repl_python()
```

```{python, echo = FALSE}
import sys
import warnings

if not sys.warnoptions:
    warnings.simplefilter("ignore")
```

# Python Basics{#Basics}

## Variables and Data Types

### Variable Assignment

```{python}
x = 5
x
# dir(x)
```


### Calculations With Variables 

```{python}
x + 2      # Sum of two variables
x - 2      # Subtraction of two variables
x*2        # Multiplication of two variables
x**2       # Exponentiation of a variable
x%2        # Remainder of a variable
x/float(2) # Division of a variable
```

### Types and Type Conversion

```{python}
type(5)        # Integer
type(5.)       # Float
type('5')      # Strings
type(5 + 3j)   # Complex Number
type(True)     # Boolean
```

## Logical Operators

```{python}
a1 = True
print(a1)
type(a1)

b1 = False
print(b1)
type(b1)

True  and True
True  and False
False and True
False and False

True or False
not True
not False

True and not False
```


## Comparison

```{python}
x1 = 30
x1 > 30
x1 == 30
x1 >= 30
x1 > 15

not x1 == 42
x1 != 42
```

## If-Else

```{python}
x2 = 30                # assign 30 to x2
if x2 > 30:            # predicate: is x2 > 30
    print("Yes")       # if True, do this
else:
    print("No")        # if False, do this
```

## Function

```{python}
def slength1(s):
    """Returns a string describing the
    length of the sequences"""
    if len(s) > 10:
        ans = 'very long'
    else:
        ans = 'normal'
    return ans


help(slength1)

slength1("Hello")
slength1("HelloHello")
slength1("Hello again")
help(dir)
dir(slength1)

def slength2(s):
    """Returns a string describing the
    length of the sequences into 
    empty, very long, normal and short"""
    if len(s) == 0:
        ans = 'empty'
    elif len(s) > 10:
        ans = 'very long'
    elif len(s) > 7:
        ans = 'normal'
    else:
        ans = 'short'
    return ans


help(slength2)
slength2("")
slength2("Good Morning")
slength2("Greetings")
slength2("Hi")
```

## Help

```{python, eval = FALSE}
# help(str)
# dir(str) 
# help(str.__add__)
# help(str.capitalize)
# dir(str.capitalize)
```


# Simple Linear Regression{#SLR}

```{python}
Income = [80, 100, 120, 140, 160, 180, 200, 220, 240, 260]
Expend = [70, 65, 90, 95, 110, 115, 120, 140, 155, 150]

import pandas as pd
# dir(pd)
# help(pd)
# help(pd.DataFrame)
# dir(pd.DataFrame)

df1 = pd.DataFrame(
 {
   "Income": Income
 , "Expend": Expend
 }
 )

print(df1)
# dir(df1)
# df1.to_html()
# df1.to_latex()
# help(df1.kurt)
# df1.kurt()
# help(df1.kurtosis)
# df1.kurtosis()
# help(df1.mean)
# df1.mean()
```

```{python}
from matplotlib import pyplot as plt
fig = plt.figure()
plt.scatter(
  x      = "Income"
, y      = "Expend"
, color  = "green"
, marker = "o"
, data   = df1
)
plt.title("Scatter plot of Weekly Income (\$) and Weekly Expenditures (\$)")
plt.xlabel("Weekly Income (\$)")
plt.ylabel("Weekly Expenditures (\$)")
plt.show()
```


```{python}
from statsmodels.formula.api import ols
from statsmodels.stats.anova import anova_lm
Reg1 = ols(formula = "Expend ~ Income", data = df1)
Fit1 = Reg1.fit()

print(Fit1.summary())
print(Fit1.params)
print(Fit1.fittedvalues)
print(Fit1.resid)
print(Fit1.bse)
print(Fit1.centered_tss)
print(anova_lm(Fit1))
```

```{python}
fig = plt.figure()
plt.scatter(
  x      = "Income"
, y      = "Expend"
, color  = "green"
, marker = "o" 
, data   = df1
)
plt.plot(df1["Income"], Fit1.fittedvalues)
plt.title("Regression plot of Weekly Income (\$) and Weekly Expenditures (\$)")
plt.xlabel("Weekly Income (\$)")
plt.ylabel("Weekly Expenditures (\$)")
plt.show()
```


# Multiple Linear Regression{#MLR}

```{python}
import numpy as np
Fertilizer = np.arange(100, 800, 100)
Rainfall   = [10, 20, 10, 30, 20, 20, 30]
Yield      = [40, 50, 50, 70, 65, 65, 80]

import pandas as pd

df2 = pd.DataFrame(
 {
   "Fertilizer": Fertilizer
 , "Rainfall": Rainfall
 , "Yield": Yield
 }
 )

print(df2)
```

```{python, eval = TRUE}
from mpl_toolkits.mplot3d import Axes3D
from matplotlib import pyplot as plt

fig = plt.figure()
ax = fig.add_subplot(111, projection = "3d")
ax.scatter(
  df2["Fertilizer"]
, df2["Rainfall"]
, df2["Yield"]
, color = "green"
, marker = "o"
, alpha  = 1
)
ax.set_xlabel("Fertilizer")
ax.set_ylabel("Rainfall")
ax.set_zlabel("Yield")
plt.show()
```


```{python}
from statsmodels.formula.api import ols
from statsmodels.stats.anova import anova_lm
Reg2 = ols(formula = "Yield ~ Fertilizer + Rainfall", data = df2)
Fit2 = Reg2.fit()
print(Fit2.summary())
print(Fit2.params)
print(Fit2.fittedvalues)
print(Fit2.resid)
print(Fit2.bse)
print(Fit2.centered_tss)
print(anova_lm(Fit2))
```

```{python, eval = TRUE}
from mpl_toolkits.mplot3d import Axes3D
from matplotlib import pyplot as plt
import numpy as np
import pandas as pd
from matplotlib import cm
fig = plt.figure()
ax = fig.add_subplot(111, projection = "3d")
ax.scatter(
  df2["Fertilizer"]
, df2["Rainfall"]
, df2["Yield"]
, color = "green"
, marker = "o"
, alpha  = 1
)
ax.set_xlabel("Fertilizer")
ax.set_ylabel("Rainfall")
ax.set_zlabel("Yield")
x_surf = np.arange(100, 720, 20)
y_surf = np.arange(10, 32, 2)
x_surf, y_surf = np.meshgrid(x_surf, y_surf)

exog = pd.core.frame.DataFrame({
   "Fertilizer": x_surf.ravel()
 , "Rainfall": y_surf.ravel()
 })
out = Fit2.predict(exog = exog)
ax.plot_surface(
           x_surf
         , y_surf
         , out.values.reshape(x_surf.shape)
         , rstride=1
         , cstride=1
         , color="None"
         , alpha = 0.4
         )
plt.show()
```

# Polynomial Regression{#PolyReg}

```{python}
Nitrogen = [0, 0, 10, 10, 20, 20]
Yield    = [5, 7, 15, 17,  9, 11]

import pandas as pd
df3 = pd.DataFrame(
 {
   "Nitrogen": Nitrogen
 , "Yield": Yield
 }
 )

print(df3)
```

```{python, eval = TRUE}
from matplotlib import pyplot as plt
fig = plt.figure()
plt.scatter(
  df3["Nitrogen"]
, df3["Yield"]
, color = "green"
, marker = "o"
)
plt.title("Scatter plot of Nitrogen and Yield")
plt.xlabel("Nitrogen")
plt.ylabel("Yield")
plt.show()
```


```{python}
from statsmodels.formula.api import ols
from statsmodels.stats.anova import anova_lm
Reg3 = ols(formula = "Yield ~ Nitrogen + I(Nitrogen**2)", data = df3)
Fit3 = Reg3.fit()
print(Fit3.summary())
print(Fit3.params)
print(Fit3.fittedvalues)
print(Fit3.resid)
print(Fit3.bse)
print(Fit3.centered_tss)
print(anova_lm(Fit3))
```

```{python, eval = TRUE}
fig = plt.figure()
plt.scatter(
  df3["Nitrogen"]
, df3["Yield"]
, color = "green"
, marker = "o"
)
plt.plot(df3["Nitrogen"], Fit3.fittedvalues)
plt.title("Regression plot of Nitrogen and Yield")
plt.xlabel("Nitrogen")
plt.ylabel("Yield")
plt.show()
```

# Regression with Dummy Variables {#RegDummy}

## Example 1

```{python}
Consumption = [5, 6, 7, 15, 16, 17]
Gender = ["Male", "Male","Male", "Female", "Female", "Female"]

import pandas as pd
df4 = pd.DataFrame(
 {
   "Consumption": Consumption
 , "Gender": Gender
 }
 )

print(df4)
```

```{python, eval = TRUE}
import pandas as pd
fig = plt.figure()
df4.boxplot(
    "Consumption"
  , by = "Gender"
  )
plt.title("Boxplot")
plt.xlabel("Gender")
plt.ylabel("Consumption")
plt.show()
```


```{python}
from statsmodels.formula.api import ols
from statsmodels.stats.anova import anova_lm
fm4 = ols(formula = "Consumption ~ Gender", data = df4)
Fit4 = fm4.fit()

print(dir(Fit4))

print(Fit4.summary())
print(Fit4.params)
print(Fit4.centered_tss)
print(anova_lm(Fit4))
```


## Example 2

```{python}
Consumption = [5, 7, 15, 17, 17, 19]
EduGroup = ["NoEdu", "NoEdu", "SchoolEdu", "SchoolEdu", "CollegeEdu", "CollegeEdu"]

import pandas as pd
df5 = pd.DataFrame(
 {
   "Consumption": Consumption
 , "EduGroup": EduGroup
 }
 )

print(df5)
```

```{python, eval = TRUE}
fig = plt.figure()
df5.boxplot(
    "Consumption"
  , by = "EduGroup"
  )
plt.title("Boxplot")
plt.xlabel("EduGroup")
plt.ylabel("Consumption")
plt.show()
```


```{python}
fm5 = ols(formula = "Consumption ~ EduGroup", data = df5)
Fit5 = fm5.fit()

print(Fit5.summary())
print(Fit5.params)
print(Fit5.centered_tss)
print(anova_lm(Fit5))
```


## Example 3

```{python}
Consumption = [51, 52, 53, 54, 56, 57, 55, 56, 58, 59, 62, 63]
Gender  = ["Male", "Male", "Male", "Male", "Male", "Male", "Female", "Female", "Female", "Female", "Female", "Female"]
Income = [80, 80, 90, 90, 100, 100, 80, 80, 90, 90, 100, 100]

import pandas as pd
df6 = pd.DataFrame(
 {
   "Consumption": Consumption
 , "Gender": Gender
 , "Income": Income
 }
 )

print(df6)
```

### Regression with same slopes and different intercepts

```{python}
from statsmodels.formula.api import ols
from statsmodels.stats.anova import anova_lm
Reg6 = ols(formula = "Consumption ~ Gender + Income", data = df6)
Fit6 = Reg6.fit()

print(Fit6.summary())
print(Fit6.params)
print(Fit6.fittedvalues)
print(Fit6.resid)
print(Fit6.bse)
print(Fit6.centered_tss)
print(anova_lm(Fit6))
```

```{python}
import matplotlib.pyplot as plt
from statsmodels.graphics.factorplots import interaction_plot
fig = plt.figure()
fig = interaction_plot(
    x        = Income
  , trace    = Gender
  , response = Fit6.fittedvalues
  , colors   = ['red','blue']
  , markers  = ['D','^']
  , xlabel   ='Income'
  , ylabel   = 'Consumption'
  )
plt.show()
```

### Regression with different slopes and different intercepts
```{python}
from statsmodels.formula.api import ols
from statsmodels.stats.anova import anova_lm
Reg7 = ols(formula = "Consumption ~ Gender*Income", data = df6)
Fit7 = Reg7.fit()

print(Fit7.summary())
print(Fit7.params)
print(Fit7.fittedvalues)
print(Fit7.resid)
print(Fit7.bse)
print(Fit7.centered_tss)
print(anova_lm(Fit7))
```

```{python}
import matplotlib.pyplot as plt
from statsmodels.graphics.factorplots import interaction_plot
fig = plt.figure()
fig = interaction_plot(
    x        = Income
  , trace    = Gender
  , response = Fit7.fittedvalues
  , colors   = ['red','blue']
  , markers  = ['D','^']
  , xlabel   ='Income'
  , ylabel   = 'Consumption'
  )
plt.show()
```
